---
title: "RNA-seq Differential Expression & Functional Enrichment Pipeline"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: cosmo
  pdf_document:
    toc: true
  word_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE, fig.width = 8, fig.height = 6)
```

# Overview

This document walks through a bulk RNA-seq workflow starting from public data on **recount3**, through differential expression testing with **DESeq2**, quality control (PCA), visualization (heatmaps, volcano plots), and functional enrichment (GO, GSEA).

Example dataset: project `SRP095512` (human, GENCODE v26 annotation).

---

# 1. Library Installation & Loading

```{r install-packages, eval=FALSE}
if (!require("BiocManager", quietly = TRUE))
  install.packages("BiocManager")

BiocManager::install(c(
  "DESeq2", "EnhancedVolcano", "tidyverse", "dplyr", "ggplot2",
  "recount3", "org.Hs.eg.db", "clusterProfiler", "AnnotationDbi",
  "pheatmap", "ComplexHeatmap", "apeglm "
))
```

```{r load-libraries}
library(DESeq2)
library(EnhancedVolcano)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(recount3)
library(org.Hs.eg.db)
library(clusterProfiler)
library(AnnotationDbi)
library(pheatmap)
library(ComplexHeatmap)
library(apeglm)
```

---

# 2. Data Acquisition (recount3)

```{r fetch-data}
rse <- recount3::create_rse_manual(
  project = "SRP095512",
  project_home = "data_sources/sra",
  organism = "human",
  annotation = "gencode_v26",
  type = "gene"
)

head(rowRanges(rse))    # genomic coordinates of sequenced genes
summary(assay(rse))     # basic min/max/mean statistics
colnames(colData(rse))  # column names of metadata
```

---

# 3. Counts Conversion

recount3 provides raw base-pair coverage, not standard read counts — convert before use in DESeq2.

```{r counts-conversion}
# finding out whether raw base-pair coverage counts are present
assayNames(rse)

# basic min/max/mean statistics pre read-count correction
summary(assay(rse))
colSums(assay(rse) / 1e6)  # counts per million

# converting raw base-pair coverage counts to read counts as a new column
assay(rse, "counts") <- transform_counts(rse)

# verifying the creation of new column
assayNames(rse)

# basic min/max/mean statistics post read-count correction
summary(assay(rse, "counts"))
colSums(assay(rse, "counts") / 1e6)  # counts per million

# creating a matrix with read counts
expr_array <- assay(rse, "counts")
dim(expr_array)
colnames(expr_array)
class(expr_array)

# storing metadata as a data frame
meta <- as.data.frame(colData(rse))
head(meta)

# setting the order level of counts as first
assays(rse) <- assays(rse)[c("counts", "raw_counts")]
assays(rse)
```

---

# 4. Building the DESeq2 Dataset

```{r build-dds}
meta$sra.sample_title

# Converting sample info into two values: Ctrl vs Pat (*varies upon experiment data)
colData(rse)$condition <- sub("\\s\\d+$", "",
                              sub("^DM\\.", "", rse$sra.sample_title))

# Viewing the count of samples in each condition
table(colData(rse)$condition)

# DESeqDataSet generation
ddSE <- DESeqDataSet(rse, design = ~condition)

colData(ddSE)$condition

# Leveling the condition factors such that control comes first
ddSE$condition <- factor(ddSE$condition, levels = c("Ctrl", "Pat"))

# verifying the level
levels(colData(ddSE)$condition)

# editing ENSEMBL rownames ENSG().1 with ENSG()
rownames(ddSE) <- sub("\\..*$", "", rownames(ddSE))
head(rownames(ddSE))
```

---

# 5. Pre-filtering Low-Count Genes

```{r prefilter}
nrow(counts(ddSE)) # checking rows prior to filtering

keep <- rowSums(counts(ddSE) > 10) >= 5  # customizable according to dataset
ddSE <- ddSE[keep, ]

nrow(counts(ddSE)) # checking rows post-filtering
```

---

# 6. Differential Expression Analysis

```{r run-deseq}
DEA <- DESeq(ddSE)

# correcting rownames
rownames(DEA) <- sub("\\..*$", "", rownames(DEA))
head(rownames(DEA))

# storing DESeq2 results
de_result <- results(DEA, alpha = 0.05)  # adjusted p-value default is 0.1
summary(de_result)  # stats about log fold change

resultsNames(DEA)

#Log fold change shrinkage for visualization and ranking
resLFC <- lfcShrink(DEA, coef="condition_Pat_vs_Ctrl", type="apeglm")
resLFC

# Reordering results based on adjusted p-value
resOrdered <- de_result[order(de_result$padj), ]  # ascending order
head(resOrdered)
tail(resOrdered)
```

---

# 7. Quality Control: PCA

Assess whether samples cluster by condition (as expected) before trusting the DE results.

```{r pca}
vsd <- vst(DEA, blind = TRUE)  # variance-stabilizing transformation

plotPCA(vsd, intgroup = "condition")
```

---

# 8. Gene ID Annotation (ENSEMBL ↔ Symbol)

```{r annotate-genes}
# editing ENSEMBL rownames ENSG().1 with ENSG()
rownames(de_result) <- sub("\\..*$", "", rownames(de_result))

# Using clusterProfiler/AnnotationDbi for fetching gene symbols
symbol <- mapIds(org.Hs.eg.db, keys = rownames(de_result),
                  column = c("SYMBOL"), keytype = "ENSEMBL")

keep <- !is.na(symbol)          # logical vector for valid gene symbol: T/F
de_result <- de_result[keep, ]  # filter out ENSEMBL IDs without valid gene symbols
symbol <- symbol[keep]          # filter symbol vector with valid gene symbols

# including a column for gene symbol
de_result$symbol <- symbol[rownames(de_result)]
head(de_result)
```

---

# 9. MA Plot

```{r ma-plot}
plotMA(de_result, ylim = c(-7, 7))


plotMA(resLFC, ylim=c(-7,7)) #MA-plot for the shrunken log2 fold changes
```

---

# 10. Heatmaps

## 10.1 Common setup

```{r heatmap-setup}
# filtering top 20 significant genes based on adjusted p-values
sig_genes <- de_result[(!is.na(de_result$padj) & de_result$padj < 0.05) & abs(de_result$log2FoldChange) > 1, ] #selecting top 20

top20 <- head(sig_genes[order(sig_genes$padj), ], 20)
top20_genes <- rownames(top20)
top20_genes

#gene expression matrix using counts and filtered genes
matrix <- counts(DEA, normalized = TRUE)[top20_genes,]
rownames(matrix) <- top20$symbol

# annotation df for sample-condition info
df <- as.data.frame(colData(ddSE)["condition"])
df
```

## 10.2 pheatmap

```{r heatmap-pheatmap}
pheatmap(matrix,
         scale = "row",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         annotation_col = df)
```

## 10.3 ComplexHeatmap

```{r heatmap-complexheatmap}
mat.z <- t(apply(matrix, 1, scale))  # scaling of gene expression matrix
colnames(mat.z) <- colnames(matrix)  # adding column names to the matrix

col_anno <- HeatmapAnnotation(df = df)

Heatmap(mat.z,
        name = "z-score",
        cluster_rows = FALSE,
        cluster_columns = TRUE,
        top_annotation = col_anno,
        row_labels = rownames(matrix),
        column_labels = colnames(mat.z))
```

---

# 11. Volcano Plots

## 11.1 ggplot2 version

```{r volcano-ggplot}
expr_df <- as.data.frame(de_result)
expr_df$significance <- abs(expr_df$log2FoldChange) > 1 & expr_df$padj < 0.05

ggplot(expr_df, aes(x = log2FoldChange, y = -log10(padj), color = significance)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("blue", "red")) +
  theme_classic()
```

## 11.2 EnhancedVolcano version

```{r volcano-enhanced}
EnhancedVolcano(expr_df,
                lab = expr_df$symbol,
                x = 'log2FoldChange',
                y = 'padj',
                legendLabels = c('NS', 'Log2 FC', 'p-adj', 'p-adj and log2 FC'),
                pCutoff = 0.05,
                FCcutoff = 1)
```

---

# 12. GO Enrichment Analysis

```{r go-enrichment}
# filtering significant ENSEMBL IDs
genes_to_test <- rownames(expr_df[abs(expr_df$log2FoldChange) > 1 & expr_df$padj < 0.05, ])

genes_to_test


# performing GO enrichment
GO_results <- enrichGO(gene = genes_to_test,
                       OrgDb = "org.Hs.eg.db",
                       keyType = "ENSEMBL",
                       ont = "BP",
                       readable = TRUE)

GO_results
```

```{r go-barplot, fig.height=8}
barplot(GO_results, showCategory = 20)
```

```{r go-dotplot, fig.height=8}
dotplot(GO_results, showCategory = 20)
```

---

# 13. Gene Set Enrichment Analysis (GSEA)

```{r gsea}
gene_list <- expr_df$log2FoldChange  # selected based on log fold change
names(gene_list) <- rownames(expr_df)
gene_list <- sort(gene_list, decreasing = TRUE)  # gseGO expects a sorted ranking

gse <- gseGO(gene_list, OrgDb = "org.Hs.eg.db", keyType = "ENSEMBL", ont = "BP")
gse_df <- as.data.frame(gse)
head(gse_df)
```

```{r gsea-plot}
gseaplot(gse, geneSetID = 1)
```

---

# Session Info

```{r session-info}
sessionInfo()
```
